---
title: "Cyclistic Bike-Share Case Study"
author: "Colton Williams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(scales)
```

## Introduction
Cyclistic’s marketing team is focused on converting casual riders into annual members.  
This analysis compares Q1 2019 and Q1 2020 trip data to identify usage patterns.

## Data Import
```{r}
df19 <- read_csv("Divvy_Trips_2019_Q1.csv")
df20 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

## Data Cleaning
```{r}
clean_2019 <- df19 %>%
  rename(started_at = start_time, ended_at = end_time,
         start_station_name = from_station_name,
         end_station_name = to_station_name,
         member_casual = usertype) %>%
  mutate(member_casual = recode(member_casual,
                                "Subscriber" = "member",
                                "Customer" = "casual"),
         started_at = ymd_hms(started_at),
         ended_at = ymd_hms(ended_at),
         ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")))

clean_2020 <- df20 %>%
  mutate(started_at = ymd_hms(started_at),
         ended_at = ymd_hms(ended_at),
         ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")))

trips <- bind_rows(clean_2019, clean_2020) %>%
  filter(ride_length > 1, ride_length < 1440) %>%
  mutate(day_of_week = wday(started_at, label = TRUE),
         month = floor_date(started_at, "month"))

# Aggregated data for charts
rides_by_day <- trips %>% count(day_of_week, member_casual, name = "rides")
avg_len_by_day <- trips %>% group_by(day_of_week, member_casual) %>% summarise(avg_ride_length_min = mean(ride_length))
monthly_volume <- trips %>% count(month, member_casual, name = "rides")
top_stations <- trips %>% count(member_casual, start_station_name, sort = TRUE)
```

## Analysis & Visualizations

### Chart 1: Rides by Day of Week
```{r chart1, fig.width=7, fig.height=5}
ggplot(rides_by_day, aes(x = day_of_week, y = rides, fill = member_casual)) +
  geom_col(position = position_dodge()) +
  labs(title = "Rides by Day of Week: Members vs Casual", 
       x = "Day of Week", 
       y = "Number of Rides") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6))
```

### Chart 2: Average Ride Length by Day
```{r chart2, fig.width=7, fig.height=5}
ggplot(avg_len_by_day, aes(x = day_of_week, y = avg_ride_length_min, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average Ride Length by Day: Members vs Casual",
       x = "Day of Week", 
       y = "Avg Ride Length (min)") +
  theme_minimal() + 
  theme(axis.text.x = element_text(size = 6))
```

### Chart 3: Monthly Ride Volume
```{r chart3, fig.width=7, fig.height=5}
ggplot(monthly_volume, aes(x = month, y = rides, color = member_casual, group = member_casual)) + 
  geom_line() + 
  geom_point() + 
  labs(title = "Monthly Ride Volume: Members vs Casual", 
       x = "Month", 
       y = "Number of Rides",
       color = "Rider Type") + 
  scale_y_continuous(labels = scales::comma_format(accuracy = 1), 
                     breaks = scales::breaks_pretty()) + 
  theme_minimal()
```

### Chart 4: Top 10 Stations by Rider Type
```{r chart4, fig.width=7, fig.height=5}
top10_stations <- top_stations %>% group_by(member_casual) %>% slice_max(n, n = 10)

ggplot(top10_stations, aes(x = reorder(start_station_name, n), y = n, fill = member_casual)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~member_casual, scales = "free") +
  labs(title = "Top 10 Start Stations: Members vs Casual",
       x = "Station", 
       y = "Number of Rides") +
  theme_minimal()
```

## Key Findings
- Members ride more on weekdays, casuals on weekends  
- Casual riders have longer ride durations  
- Members favor commuter stations; casuals favor tourist stations  

## Recommendations
- Weekend promos to convert casual riders  
- Employer partnerships for memberships  
- QR-code promos at top casual stations  
